Index measures for oak decline severity using phenotypic descriptors

### 2019-12-10 manuscript/manuscript.Rmd
added to introduction and methods

- Increasing importance of monitoring 
- Acute and chronic decline associated phenotypes
- The oak decline spectrum move away from simplistic binary categorisations
- Manual assignment of decline status difficult prone to error for due to contribution of many factors. 
- Random forest
- Unsupervised learning for complex patterns between descriptors.

Oak declines are complex syndromes that are facilitated by a variety of both abiotic and biotic factors.
Abiotic factors can include both predisposing and inciting factors such as drought and are often

Biotic factors can include...

As a results of this, there are many visual phenotypic descriptors encompassing the entirety of an oak tree that contribute to what can be defined as a declining oak tree.
A severely declining oak is most prominently characterised by very poor crown condition and numerous dead branches (ref).

Compounding this complexity is the addition of different types of oak decline (AOD) that are often categorised based on the duration over which a tree has been declining or the biotic agents/factors associated with a declining tree.
For instance a chronically declining oak can subsequently begin to also develop the stem bleeds associated with AOD.
Accurately diagnosing in the field the severity and type of oak decline present is difficult and can requires much experience, expertise and training.

@@

Here, we propose a framework to enable quantitative visible oak decline severity measurement that utilises a range of targeted phenotypic descriptors and unsupervised random forest machine learning.
These measures include:

@@

- Site descriptions
- Field phenotyping protocol
- Crown assessments
- Cardinal assessments
- Stem assessments - 2m base, 3m base

Phenotypic data were collected from a total of 174 trees from 9 sites from across the south, east and west of England (Figure #).
A total of 30 phenotypic descriptors were collected relating to the trees shape, size, crown condition and the presence of biotic agents including insects, bacteria and fungi.

@@

- Aggregation of cardinal observations.
- Random forest and MDS analyses
- Descriptions of collected phenotypic descriptors and manual status assignment

@@

Cardinal crown radius measurements were averaged.

@@

The code used to generate the decline indexes is available as an R package at [https://github.com/jasenfinch/pdi].
All raw data and code used to generate this manuscript are available at [https://github.com/jasenfinch/Index_measures_for_oak_decline_severity_using_phenotypic_descriptors]

@@

**Estimated crown volume (m^3^)** - estimated volume of crown as an ellipsoid proportional to the crown condition.

$$v = \frac{c(h - l)}{150}\pi r^2$$

@@

**Estimated bleed prevalence (%)** - estimated percentage of the surveyed trunk area affected by bleeds or black staining.

@@

***Agrilus* exit hole density (m^-2^)** - density of observed *Agrilus* exit holes across the surveyed trunk area.
$$e = \frac{h}{\frac{2d}{100}\pi}$$
$h$ = number of *Agrilus* exit holes, $d$ = diameter at breast height (cm)

@@

- Accessability of collecting descriptors
- Suitability of collected phenotypic descriptors for describing decline status
- Concession of model interpretability to predictive capability by using a machine learning approach
- Trends of key phenotypic descriptors withing the PDI and DAI measures
- Sensitivity provided for integration with large scale omics analyses as well as long term monitoring for forest condition
- Index thresholds for category definitions (healthy, declining, severe decline etc.)
- When viewed in the 2D continuum, the categorisation of AOD to only include bleeds of certain bacteria seems likely to be too specific and by definition should only describe the period of time under which the tree have been declining.
Perhaps subcategories of AOD would be more suitable such as Agrilus associated AOD or Armillaria associated AOD.

>>>>

- Monitoring currently focusses on crown assessments
- Acute and chronic decline associated phenotypes
- The oak decline spectrum move away from simplistic binary categorisations
- Manual assignment of decline status difficult prone to error for due to contribution of many factors. 
- Unsupervised learning for complex patterns between descriptors.

Oak declines are complex syndromes that are facilitated by a variety of both abiotic and biotic factors.
Abiotic factors can include both predisposing and inciting factors such as drought and are often
Biotic agents can include insect infestation such as the larvae of bark boring bettle *Agrilus biguttatus* associated

As a results of this, there are many visual phenotypic descriptors encompassing the entirety of an oak tree that contribute to what can be defined as a declining oak tree.
A severely declining oak is most prominently characterised by very poor crown condition and numerous dead branches (ref).
It is these characteristics that many studies use as the primary indicator of tree decline severity [@zarnoch2004].

Compounding this complexity is the addition of different types of oak decline (AOD) that are often categorised based on the duration over which a tree has been declining or the biotic agents/factors associated with a declining tree.
For instance a chronically declining oak can subsequently begin to also develop the stem bleeds associated with AOD [@denman2014].
Accurately diagnosing in the field the severity and type of oak decline present is difficult and can requires much experience, expertise and training.

It is becoming increasingly important that oak tree health within woodlands is accurately and sensitively monitoried to enable the detection of decline as well as it's spread and severity over time to enable accurate forecasting and inform forestry management practices (ref).
Monitoring methods need to take advantage of modern data analysis techniques whilst still retaining practical simplicity in the field.

@@

Here, we propose a framework to enable quantitative visible oak decline severity measurement that utilises a range of targeted phenotypic descriptors and unsupervised random forest machine learning.
The two resulting measures are:

@@

- Site types
- Descriptor types
- Field phenotyping protocol
- Crown assessments
- Stem assessments - 2m base, 3m base

Phenotypic data were collected from a total of 174 trees from 9 sites from across the south, east and west of England.

A total of 30 phenotypic descriptors were collected relating to the trees shape, size, crown condition and the presence of biotic agents including insects, bacteria and fungi.
These descriptors are shown in Table #.
Crown radius, canopy closure and tap tests for dead stem tissue were measured for each cardinal point.
AOD stem symptoms (active bleeds, black staining, callused wounds) were collected from the basel 3m of the stem.
Each individual

@@

- Descriptors converted to presence and absence
- Aggregation of cardinal observations.
- Symptom type sizes average and frequency taken
- Random forest and MDS analyses
- Descriptions of collected phenotypic descriptors and manual status assignment

@@

Cardinal assessments were averaged to give a single value.

@@

The code used to generate the decline indexes is available as an R package pdi (v) at https://github.com/jasenfinch/pdi.
All raw data and code used to generate this manuscript are available at https://github.com/jasenfinch/Index_measures_for_oak_decline_severity_using_phenotypic_descriptors.

@@

**Estimated crown volume (m^3^)** - estimated volume of crown as an paraboloid proportional to the crown condition. Based on the composite crown volumne from @zarnoch2004 .

$$v = \frac{\pi r^2c(h-l)}{200}$$

@@

**Estimated bleed prevalence (%)** - estimated percentage of the surveyed 3m basal trunk area affected by bleeds or black staining.

@@

***Agrilus* exit hole density (m^-2^)** - density of observed *Agrilus* exit holes across the surveyed 2m basal trunk area.
$$e = \frac{h}{\frac{2d}{100}\pi}$$
$h$ = number of *Agrilus* exit holes, $d$ = diameter at breast height (cm)

@@

- Accessability of collecting descriptors
- Suitability of collected phenotypic descriptors for describing decline status
- Concession of model interpretability to predictive capability by using a machine learning approach
- Trends of key phenotypic descriptors withing the PDI and DAI measures
- Sensitivity provided for integration with large scale omics analyses as well as long term monitoring for forest condition
- Index thresholds for category definitions (healthy, declining, severe decline etc.)
- Semi-supervised can be used if unsupervised model is unable to pull out decline severity scale
- When viewed in the 2D continuum, the categorisation of AOD to only include bleeds of certain bacteria seems likely to be too specific and by definition should only describe the period of time under which the tree have been declining.
Perhaps subcategories of AOD would be more suitable such as Agrilus associated AOD or Armillaria associated AOD.
